---
title: "Assignment 3: Data Exploration"
author: "Christopher Starr"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}

#Read packages
library(tidyverse); library(lubridate); library(here); library(ggplot2)

#Read in data, don't convert strings to factors
Neonics <- read.csv(here('Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'), stringsAsFactors = TRUE)
# reading in as .csv
str(Neonics)
#taking a look at the datase

Litter <- read.csv(here("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"), stringsAsFactors = TRUE)
# reading in as .csv
str(Litter)
#taking a look at the datase
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Likely we are looking for links between populations of certain insects and use of insecticides. Ideally we want to find a strong correlation between declines and use of insectices so we can understand what insects are being killed by their use and how dramatically the population is harmed. Maybe then we can match this information with what we know about different classes of affected insects and thier role in other areas of plant/animal life.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Likely we want to understand what trees and other forest plants are dropping to understand if their overall health is declining. For instance are they dropping more leaves and branches than before? Is there a decline in seed or other "reproductive" litter count or density?

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. The team set ground-level and above ground traps to collect material in areas with vegation that is more than 2 meters tall. They meausured the materials at different times depending on the collection method (above ground every 1-2 weeks, on ground once per year).
 2. The ground and above-ground traps were paired together and were either placed randomly (in thick forest areas) or targetted (where the forest was not as thick). The sampling occurred in something called Tower Plots.
 3. They measured up to .01g accuracy 8 types of vegetation (Leaves, needles, seeds, etc.). 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
str(Neonics)
# Taking a look at the Neonics dataset
# 4632 obersavtions with 30 variables each
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
sort(summary(Neonics$Effect), decreasing = TRUE)
# asing R to put the data in order for the Neonics file, Effect vector, decreasing by 
# frequency. 
```

> Answer:Population (1803), Mortality (1493), and Behavior (360) are the largest categories.
By measuring the litter we can see if there are more types of litter so we can understand
if the population of a certain type of litter is increasing or decreasing. We can also check
mortality, which might be checking to see if there are fully dead trees in the sample. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
sort(summary(Neonics$Species.Common.Name), decreasing = TRUE)
# asing R to put the data in order for the Neonics file, Species.Common.Name vector, decreasing by 
# frequency. 
summary(Neonics$Species.Common.Name, maxsum = 10)
# asking R to show me the ten largest categories.
```

> Answer: Most of the top 10 is made up of bee species. Bees are thought to have an outsized impact on ecosystems and pollenation. The other names sound like paracites of interest. Perhaps their numbers are being closely examed to understand how bad their presence is in the area(s) being studied. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
class(Neonics$Conc.1..Author.)
# asking R to tell me the class or Conc.1..Author. 
```

> Answer: The vector is a Factor data type. It is not a number because there are a lot of special characters in the vector list (< and >, /, etc) which cannot be processed as numbers. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics, aes(x = Publication.Year)) +
  geom_freqpoly()
# asking R to generate a graph showing x as years and y as number of publications per year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly(binwidth = 1)
# asking R to generate a graph showing years/studies per year with a different color line
# to show different locations of where the data was gathered.
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: In the early 90s it was a close match between Field natural and Lab. In the late 90s Field natural became dominant. Throughout the 2000s it has been most Lab as the dominant location with the exception of 2009 when Field natural had a big spike.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics, aes(x=Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
# asking R to create a bar chart of Endpoints to show frequency. Adding theme elements
# to help with aesthetics since the data is not easily visible without some adjustments.

```

> Answer: the most common endpoints are NOEL and LOEL. 
No-observable-effect-level: highest dose (concentration)
producing effects not significantly different from responses of
controls according to author's reported statistical test
(NOEAL/NOEC)
Lowest Observed Effects Residue: The lowest residue
concentration producing effects that were significantly different
from responses of controls according to author's reported
statistical test

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
# asking R what collectDate is. Initial Response = "factor"

Litter$collectDate <- ymd(Litter$collectDate)
#asking R to use lubridate to to convert the column to date format.

class(Litter$collectDate)
#verifying the class changed. result = "Date"
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
#asking R to tell me how many plotID values there are
sort(summary(Litter$plotID), decreasing = TRUE)
#asking R to show me the list of all plotIDs and their count, in descending order.

```

> Answer:Unique returns a count of how many different values are in the vector and lists them. Summary returns the list of each vector AND the number of times each value occurs in the vector. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
