---
title: "Assignment 5: Data Visualization"
author: "Christopher Starr"
date: "Spring 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

#loading in libraries my code will use
library(tidyverse);library(lubridate);library(here)
library(ggridges); library(cowplot)

#checking where my code is looking around in the directory
here()

#Reading in the datasets I will be using
LakeData <-
read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

ForestLitterData <-
read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)

#2 

#checking date formats
class(LakeData$sampledate)
class(ForestLitterData$collectDate)

#dates are showing as factors, using lubridate to make them date format
LakeData$sampledate <- ymd(LakeData$sampledate)
ForestLitterData$collectDate <- ymd(ForestLitterData$collectDate)

#checking date formats
class(LakeData$sampledate)
class(ForestLitterData$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

# Create custom theme with a Minnesota Vikings color scheme - purple and gold elements
mytheme <- theme(
    # Plot background
    panel.background = element_rect(fill = "#686868", color = "#9370DB"),
    plot.background = element_rect(fill = "#787878"),
    
    # Axis text and titles - changed axis titles to black
    axis.text = element_text(color = "gold", family = "mono", size = 11),
    axis.title = element_text(color = "black", family = "mono", face = "bold", size = 12),  # Changed to black
    
    panel.grid.major = element_line(color = "#9370DB", linetype = "dashed", linewidth = 0.2),
    panel.grid.minor = element_line(color = "#9370DB", linetype = "dotted", linewidth = 0.1),
    
    legend.background = element_rect(fill = "#787878"),
    legend.text = element_text(color = "gold", family = "mono"),
    legend.title = element_text(color = "#9370DB", family = "mono", face = "bold"),
    legend.position = "top"
)

# Set as default theme for future plots
theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
phos_phorusbyphate <-
  ggplot(LakeData, aes(x = tp_ug, y = po4)) + 
  geom_point() +
  xlim(0, 100) +
  ylim(0, 50)+
  geom_smooth(
    method = lm,
    se=FALSE
  )

#Print the plot
print(phos_phorusbyphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

#create a factor for months in the correct order
month_labels <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                 "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
LakeData$month <- factor(LakeData$month, 
                        levels = 1:12,
                        labels = month_labels)

# Temperature boxplot (with legend)
temp_plot <- ggplot(LakeData, aes(x = month, y = temperature_C, fill = lakename)) +
  geom_boxplot() +
  labs(x = element_blank(),
       y = "Temp.\n(°C)",
       fill = "Lake") +
  theme(axis.text.x = element_blank())

print(temp_plot)

# Total Phosphorus boxplot
tp_plot <- ggplot(LakeData, aes(x = month, y = tp_ug, fill = lakename)) +
  geom_boxplot() +
  labs(x = element_blank(),
       y = "Total P\n(µg/L)") +
  theme(legend.position = "none",
        axis.text.x = element_blank())

print(tp_plot)

# Total Nitrogen boxplot (now with month labels)
tn_plot <- ggplot(LakeData, aes(x = month, y = tn_ug, fill = lakename)) +
  geom_boxplot() +
  labs(x = "Month",
       y = "Total N\n(µg/L)") +
  theme(legend.position = "none")

print(tn_plot)

# Combine plots

combined_plot <- plot_grid(
  temp_plot,
  tp_plot,
  tn_plot,
  ncol = 1,
  align = "v",
  rel_heights = c(1, 1, 1.2)
)

print(combined_plot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: We don't see a lot of measures of Nitrogen or Phosphorus in Jan/Feb/Mar/Apr/Oct/Nov/Dec. The temperatures follow a similar range for both lakes but the phosphorus jumps quite a bit higher in Peter lake, as does the Nitrogen

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
#create a plot that looks at the needle weight by data
needles_plot <- ForestLitterData %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(
    mapping = aes(
      x = collectDate, 
      y = dryMass,
      color = nlcdClass)) + #color dots by nlcdClass
  geom_point()+
  theme(legend.title = element_blank())

#display the plot
print(needles_plot)

#7
#create a plot that looks at the needle weight by data
needles_plot_facets <- ForestLitterData %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(
    mapping = aes(
      x = collectDate, 
      y = dryMass,
      color = nlcdClass)) +
  geom_point()+
  facet_wrap(facet=vars(nlcdClass))

#display the plot
print(needles_plot_facets)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think #7 is more effective. It more clearly allows for the viewer to see the weight and clustering across different areas. 