---
title: "Assignment 9: Spatial Analysis in R"
author: "Christopher Starr"
output: html_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single **HTML** file.


## DATA WRANGLING

### Set up your session

1.    Import libraries: tidyverse, sf, leaflet, here, and mapview
2.    Execute the `here()` command to display the current project directory

```{r Prepare workspace, error=F, warning=F}
#1. 
#install.packages(c("tidyverse"),("sf"),("leaflet"))


# Load the libraries
library(tidyverse)
library(sf)
library(leaflet)
library(here)
library(mapview)

#2. 
# Display the current project directory
here()

```


### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3.    Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4.    Reveal the dataset's coordinate reference system
5.    Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties

ne_counties <- st_read(here("Data/Raw/Spatial/cb_2018_us_county_20m.shp")) %>%
  filter(STATEFP == "31") # Nebraska's FIPS code is 31

#4. Reveal the CRS of the counties features
st_crs(ne_counties)

#5. Plot the data
#Using mapview (interactive)
mapview(ne_counties)

#Using ggplot2
ggplot(data = ne_counties) +
  geom_sf() +
  theme_minimal() +
  ggtitle("Nebraska Counties")

```

6.    What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (In other words, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: lookup the EPSG code on <https://epsg.io> or <https://spatialreference.org>)

> ANSWER: The CRS is EPSG:4269 -- is a geographic coordinate system (using angular units of degrees).This CRS is associated with the North American Datum 1983 (NAD83)


### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7.    Read the `NWIS_SiteInfo_NE_RAW.csv` file into a standard dataframe, being sure to set the `site_no` field as well as other character columns as a factor. 

8.    Display the structure of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe
library(readr)
library(dplyr)

gage_locations <- read_csv(here("Data/Raw/NWIS_SiteInfo_NE_RAW.csv")) %>%
  mutate(across(where(is.character), as.factor))

# 8. Display the structure of this dataset
str(gage_locations)

# Also display column names
names(gage_locations)


```

9.    What columns in the dataset contain the x and y coordinate values, respectively?  
> ANSWER: When plotting with sf the order is (longitude, latitude) - which is (x, y) - not (latitude, longitude) as many people intuitively expect. So x=dec_lat_va and y=dec_long_va.

### Convert the dataframe to a spatial features ("sf") dataframe

10.   Convert the dataframe to an sf dataframe.
    *   Note: These data use the same coordinate reference system as the counties dataset

11. Display the structure of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
# 10. Convert the dataframe to an sf dataframe
gage_locations_sf <- st_as_sf(gage_locations, 
                             coords = c("dec_long_va", "dec_lat_va"),
                             crs = 4269)  # NAD83 EPSG code we identified earlier

# 11. Display the structure of the resulting sf dataframe
str(gage_locations_sf)


```

12.   What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: New field: geometry; Disappeared fields: dec_lat_va, dec_long_va
>


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
  *   Be sure the datasets are displayed in different colors
  *   Title your plot "NWIS Gage Locations in Nebraska"
  *   Subtitle your plot with your name

```{r Plot the spatial features}
#13. 
ggplot() +
  # Add county boundaries
  geom_sf(data = ne_counties, fill = "lightblue", color = "darkgray") +
  # Add gage locations
  geom_sf(data = gage_locations_sf, color = "red", size = 2) +
  # Add title and subtitle
  labs(title = "NWIS Gage Locations in Nebraska",
       subtitle = "Christopher Starr") +
  # Add a clean theme
  theme_minimal()

```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe
    *   Pay attention to which fields should be imported as factors!
15. Show the structure of the dataframe.
16. Join our site information (already imported above) to these gage height data
    *   The `site_no` and `station_nm` can both/either serve as joining attributes
    *   Construct this join so that the result only includes records features where both tables have data (N=136)
17. Show the column names of this resulting spatial dataframe
18. Show the dimensions of the resulting joined dataframe

```{r Read in gage site data}
# 14. Read the NWIS_SiteFlowData_NE_RAW.csv dataset as a dataframe
# Setting character fields as factors
flow_data <- read_csv(here("Data/Raw/NWIS_SiteFlowData_NE_RAW.csv")) %>%
  mutate(
    # Convert site_no and station_nm to factors
    site_no = as.factor(site_no),
    station_nm = as.factor(station_nm),
    # Parse the date column with the correct format (ymd_hms)
    date = ymd_hms(date)
  )

# 15. Show the structure of the dataframe
str(flow_data)

# 16. Join the flow data to our NWIS gage location spatial dataframe
# Using an inner join to include only records where both tables have data
gage_locations_with_flow <- gage_locations_sf %>%
  inner_join(flow_data, by = "site_no")

# 17. Show column names of the resulting spatial dataframe
names(gage_locations_with_flow)

# 18. Show the dimensions of the joined dataframe
dim(gage_locations_with_flow)

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.
19.   Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
    *   Show the magnitude of gage height by color, shape, other visualization technique.

```{r map gage sites}
#Map the points, sized by gage height
# Transform data to WGS84 (EPSG:4326)
ne_counties_wgs84 <- st_transform(ne_counties, 4326)
gage_locations_flow_wgs84 <- st_transform(gage_locations_with_flow, 4326)

# Create a color palette based on gage height
pal <- colorNumeric(
  palette = "YlOrRd",  # Yellow-Orange-Red palette
  domain = gage_locations_flow_wgs84$gage_ht
)

# Create the leaflet map
leaflet() %>% 
  # Add base tiles (default OpenStreetMap)
  addTiles() %>%  
  # Add county polygons
  addPolygons(
    data = ne_counties_wgs84, 
    color = "orange", 
    weight = 1, 
    smoothFactor = 0.5,   
    opacity = 1.0, 
    fillOpacity = 0.2,
    fillColor = "white"
  ) %>% 
  # Add gage locations as circle markers
  addCircleMarkers(
    data = gage_locations_flow_wgs84,
    radius = ~gage_ht/2,  # Size proportional to gage height
    color = "black",
    weight = 1,
    fillColor = ~pal(gage_ht),
    fillOpacity = 0.8,
    popup = ~paste0(
      "<b>", station_nm.x, "</b><br>",
      "Gage Height: ", gage_ht, " meters<br>",
      "Date: ", format(date, "%Y-%m-%d %H:%M"), "<br>",
      "Site Number: ", site_no
    )
  ) %>%
  # Add a legend
  addLegend(
    position = "bottomright",
    pal = pal,
    values = gage_locations_flow_wgs84$gage_ht,
    title = "Gage Height (m)",
    opacity = 0.7
  )

```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r spatial analysis}
#20 Transform the counties and gage location datasets to UTM Zone 14N
ne_counties_utm <- st_transform(ne_counties, 32614)
gage_locations_flow_utm <- st_transform(gage_locations_with_flow, 32614)


#21 Plot the data
county_map <- mapview(ne_counties_utm, 
                     alpha.regions = 0.3,
                     color = "gray", 
                     legend = FALSE)

# For gage locations - sized by gage height
gage_map <- mapview(gage_locations_flow_utm, 
                   zcol = "gage_ht",
                   cex = "gage_ht",  # Size by gage height
                   cex.max = 10,     # Maximum size of points
                   alpha = 0.8,
                   layer.name = "Gage Height (m)")

# Combine the maps (this will show both layers together)
county_map + gage_map


```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Saunders county from your projected county sf dataframe 
23. Select the gage sites falling within that county to a new spatial dataframe
24. Select the gage sites within 15km of the county to a new spatial dataframe
25. Create a plot showing (each symbolized distinctly):
 * all Nebraska counties,
 * the selected county, 
 * the gage sites in that county,
 * and the gage sites within 15 km of the county


```{r spatial selection}
#22. Selected Sounder country
saunders_county <- ne_counties_utm %>%
  filter(NAME == "Saunders")

# 23. Select the gage sites falling within that county
gages_in_county <- gage_locations_flow_utm %>%
  st_filter(saunders_county, .predicate = st_intersects)

# 24. Create a 15km buffer around Saunders county
saunders_buffer <- st_buffer(saunders_county, dist = 15000)  # 15000 meters = 15km

# Then select gages that intersect this buffer but are not in the county
gages_near_county <- gage_locations_flow_utm %>%
  st_filter(saunders_buffer, .predicate = st_intersects) %>%
  filter(!site_no %in% gages_in_county$site_no)  # Exclude gages already in the county

# 25. Create map with mapview
# All Nebraska counties (light gray)
all_counties <- mapview(ne_counties_utm, 
                       alpha.regions = 0.1, 
                       color = "gray", 
                       legend = FALSE)

# Selected county (Saunders - orange highlight)
selected_county <- mapview(saunders_county, 
                          color = "navy",       # Dark blue border 
                          col.regions = "#8888FF",  # Light blue fill
                          alpha.regions = 0.4, 
                          layer.name = "Saunders County")

# 15km buffer (blue outline)
county_buffer <- mapview(saunders_buffer, 
                        alpha.regions = 0, 
                        color = "blue", 
                        lwd = 1.5,
                        layer.name = "15km Buffer")

# Gages in county (green)
gages_in <- mapview(gages_in_county, 
                   col.regions = "darkgreen", 
                   cex = 5, 
                   layer.name = "Gages in County")

# Gages near county (purple)
gages_near <- mapview(gages_near_county, 
                     col.regions = "purple", 
                     cex = 4, 
                     layer.name = "Gages within 15km")

# Combine all layers
all_counties + selected_county + county_buffer + gages_in + gages_near

```
